Eddy covariance towers as sentinels of abnormal radioactive material releases

### 2024-08-19 manuscript/manuscript.tex
text edits

\section*{Abstract}	
Ensuring accurate detection and attribution of abnormal releases of radioactive material is critical for protecting human health and safety. Most commonly, such detection is accomplished via active monitoring approaches involving the collection of physical samples. This is labor intensive and limits the temporal and spatial resolution of any detected events to a relatively coarse level. As an alternative first step towards passive monitoring, we developed an approach using eddy flux tower data records to identify signals from a known abnormal release and quantify the extent to which that signal also occurs at other times in the data record. Through two case-studies, one of which targeted the Fukushima nuclear disaster and the other targeting a radioisotope production facility, we tested our approach and identified several potential heretofore unidentified abnormal releases that were consistent with atmospheric circulation patterns and/or wind direction from known release sites. Because our approach is relatively simple and is resistant to systematic errors in the observational record, it has broad applicability beyond specific constituents and ecosystem types to identify a wide variety of limited-duration anomalies in flux tower data to ensure human health and industrial safety.

@@

We specifically focused on short-term responses of vegetation (on the order of days to weeks) to potential deposition of radioactive material rather than chronic long-term signals (on the order of years to decades). In the first case study, we tested the ability of various data mining techniques to recover the signal from the closest tower (approximately 18 km away) to a known release coming from the Institut des Radioelements (IRE) in Belgium (hereafter called the known release location or the “facility”) nominally occurring on August 23, 2008. The IRE release originated from waste tanks into the atmosphere and lasted for several days totaling 50 GBq 131I \citep{carle2010individual}. For example, grass samples collected in the vicinity of the facility found radioactivity levels of 5000 Bq/kg. In the second case study, we further explored potential signals influx tower data related to the Fukushima Nuclear Disaster nominally occurring on March 11, 2011 releasing a massive radioactive plume that reached North America in 5 days eventually dispersing throughout the entire northern hemisphere \citep{meszarosPredictabilityDispersionFukushimaderived2016}. We used the signal identified during the known release to identify additional potential abnormal releases occurring at other times in the data record and at other nearby eddy flux towers. While little is known about the direct or indirect effects of radioactive material on vegetation, we expect that exposure may cause declines in photosynthetic $CO_2$ uptake. This may be expressed as a change in the slope of the $CO_2$ relationship with other environmental variables (e.g. air temperature, humidity, precipitation, etc.) before, during, and after exposure.

@@

For the IRE case study, we gathered measurements from eddy covariance platforms (i.e., “flux towers”) located in Belgium, which are part of the ICOS (Integrated Carbon Observation System) network available through the European Fluxes Database (http://www.europe-fluxdata.eu/). Only three sites in proximity to the known release location had sufficient long-term data for our anticipated data mining efforts (BE-Lon, BE-Bra, and BE-Vie sites, Figure 1). Each of the three flux tower sites are located within a crop ecosystem type and had almost 10 years of data available from 2004 through 2013 recorded at 30 min intervals. While BE-Lon had a near-continuous data record during this period, BE-Bra had a long period of missing data in 2005, and both BE-Bra and BE-Vie had a long period of missing data in 2009. The distance between IRE and the flux tower sites are approximately 19 km, 95 km, and 105 km for the BE-Lon, BE-Bra, and BE-Vie respectively.

For the Fukushima case study, we gathered measurements from flux towers located in the Northern Hemisphere (Western United States), which are part of the Ameriflux network (https://ameriflux.lbl.gov/). For contrast to the Northern Hemisphere sites where we expect a higher likelihood of impact, we also gathered measurements from flux towers located in the Southern Hemisphere, which are part of the OzFlux network \citep{cleverly2011alice}. We selected sites with sufficient long-term data to cover the 2011 date of the disaster and excluded boreal sites with extreme snow cover related seasonality. We ultimately selected three sites (US-Wrc, US-GLE, and OZ-Mul, Figure 1B).

@@

This interaction model is fit for a window of time encompassing the known release event and for every other possible window in the period-of-record. The overall length of the window was treated as an adjustable hyperparameter since the length of time for an abnormal material release to reach the footprint of the flux tower may be event-dependent and/or unknown. In addition, the length of time for deposited material to induce a possible vegetation response (or at least a flux tower anomaly) is an unknown quantity. Thus, in the default base case, we set the window length (n\_days) as 7 days, but we ran hyperparameter experiments to determine the result of setting it to longer periods of 10 and 14 days. The window length affects the duration of time designated as before and after the event. The period of time designated as during the event was set as 2 days in all cases. We compared the effect size of the interaction term ($\beta$3) during the event-encompassing window against all other windows in a “rolling” analysis. Therefore, the overall analysis compares the interaction term of the window of the known event (which occurred on August 23, 2008) against a window centered on every other observation. Specific windows that had an effect size at least as large as that of the known event and with wind conditions defined as “towards” the flux tower were flagged as “event detections”. The frequency of these event detections are reported as “event detections”.

@@

For the IRE case study, among all possible pairwise combinations of dependent and independent variables (n=51), we found that only the $CO_2$ versus air temperature comparison had any measure of predictability from the model ($R^2$ > 0.05), strong event effect size, and wind direction coming from the release location to towers in ICOS (Table 1). Although other variable pairs such as sensible heat flux and net radiation had a stronger effect size for the BE-Lon site, they had a weak linear relationship across the other sites. When we focus on the period of the abnormal release event in particular rather than the overall period of record, we found that $CO_2$ versus air temperature relationship at each of the three focal towers had similar $R^2$ values and effect sizes (Table 1). Figure 2 provides a visual example of the effect size framing in our analysis. Note how the apparent interaction effect between time periods around the abnormal release event in Figure 2A (before, during, after) translates to a relatively high effect size (Figure 2C) whereas in Figure 2B at an arbitrary time point, there is no apparent interaction effect. This corresponds to a low effect size in Figure 2D.

@@

Another contributor to the success of our approach was the fact that Belgium has a relatively dense network of flux towers with long data records. As a result, we were able to compare event detections at the closest tower against more distant towers (refer to Figure 1), increasing our confidence that we were seeing a true signal and not spurious noise. This level of flux tower density may be found in the United States, Western/Central Europe, Japan, and Australia, which have high densities of flux towers, but not throughout the rest of the world \citep{baldocchi2001fluxnet,pastorello2020fluxnet2015}. As a result, this may impose geographical limits on our approach. Another attribute that likely aided our investigations was the fact that all three of the flux towers we examined were located in similar agricultural ecosystem types. Although investigating whether or not flux tower combinations with differing ecosystems would yield similar results is beyond the scope of the present study, we suspect that the data records among towers in disparate ecosystem types would differ too drastically to be of comparative use. For example, \citet{pastorello2020fluxnet2015} show that towers in the crop ecosystem type, of which all three sites used in this study belong, have a relatively narrow distribution of fluxes (i.e., gross primary production, GPP) compared to towers in the grassland or evergreen broadleaf forest types. One reason for this narrow spread in crop ecosystems may be the relative homogeneity of flux tower footprints in crop ecosystems, which typically extend to within 1000m of the tower depending on atmospheric conditions \citep[e.g., wind speed and direction,][]{chuRepresentativenessEddyCovarianceFlux2021}. Given that towers in other ecosystem types with a more heterogeneous flux tower footprint have a wider distribution of fluxes \citep[e.g.][]{pastorello2020fluxnet2015}, we suspect that having all the towers in our study located in a homogenous ecosystem type was helpful in reducing the spikiness and spread of the data.

>>>>

\section*{Abstract}	
Ensuring accurate detection and attribution of abnormal releases of radioactive material is critical for protecting human health and safety. Most commonly, such detection is accomplished via active monitoring approaches involving the collection of physical samples. This is labor intensive and limits the temporal and spatial resolution of any detected events to a relatively coarse level. As an alternative first step towards passive monitoring, we developed an approach using eddy flux tower data records to identify signals from a known abnormal release and quantify the extent to which that signal also occurs at other times in the data record. Through two case-studies, one of which targeted the Fukushima nuclear disaster and the other targeting an abnormal release event at a radioisotope production facility in Fleurus Belgium, we tested our approach and identified several potential heretofore unidentified abnormal releases that were consistent with atmospheric circulation patterns and/or wind direction from known release sites.Because our approach is relatively simple and is resistant to systematic errors in the observational record, it has broad applicability beyond specific constituents and ecosystem types to identify a wide variety of limited-duration anomalies in flux tower data to ensure human health and industrial safety.

@@

We specifically focused on short-term responses of vegetation (on the order of days to weeks) to potential deposition of radioactive material rather than chronic long-term signals (on the order of years to decades). In the first case study, we tested the ability of various data mining techniques to recover the signal from the closest tower (approximately 18 km away) to a known release coming from the Institut des Radioelements (IRE) in Belgium (hereafter called the known release location or the “facility”) nominally occurring on August 23, 2008. The IRE release originated from waste tanks into the atmosphere and lasted for several days totaling 50 GBq 131I \citep{carle2010individual}. For example, grass samples collected in the vicinity of the facility found radioactivity levels of 5000 Bq/kg. In the second case study, we further explored potential signals influx tower data related to the Fukushima Nuclear Disaster nominally occurring on March 11, 2011 releasing a massive radioactive plume that reached North America in 5 days that eventually dispersed throughout the entire northern hemisphere \citep{meszarosPredictabilityDispersionFukushimaderived2016}. We used the signal identified during the known release to identify additional potential abnormal releases occurring at other times in the data record and at other nearby eddy flux towers. While the body of literature on direct or indirect effects of radioactive material on vegetation is limited, we expect that exposure may cause declines in photosynthetic $CO_2$ uptake. This may be expressed as a change in the slope of the $CO_2$ relationship with other environmental variables (e.g. air temperature, humidity, precipitation, etc.) before, during, and after exposure.

@@

For the IRE case study, we gathered measurements from eddy covariance platforms (i.e., “flux towers”) located in Belgium, which are part of the ICOS (Integrated Carbon Observation System) network available through the European Fluxes Database (http://www.europe-fluxdata.eu/). Three sites in proximity to the known release location had sufficient long-term data for our anticipated data mining efforts (BE-Lon, BE-Bra, and BE-Vie sites, Figure 1). Each of the three flux tower sites are located within a crop ecosystem type and had almost 10 years of data available from 2004 through 2013 recorded at 30 min intervals. While BE-Lon had a near-continuous data record during this period, BE-Bra had a long period of missing data in 2005, and both BE-Bra and BE-Vie had a long period of missing data in 2009. The distance between IRE and the flux tower sites are approximately 19 km, 95 km, and 105 km for the BE-Lon, BE-Bra, and BE-Vie respectively.

For the Fukushima case study, due to the direction of the plume, we gathered measurements from flux towers located in the Northern Hemisphere (Western United States), which are part of the Ameriflux network (https://ameriflux.lbl.gov/). For contrast to the Northern Hemisphere sites where we expect a higher likelihood of impact, we also gathered measurements from flux towers located in the Southern Hemisphere, which are part of the OzFlux network \citep{cleverly2011alice}. We selected sites with sufficient long-term data to cover the 2011 date of the disaster and excluded boreal sites with extreme snow cover related seasonality. We ultimately selected three sites (US-Wrc, US-GLE, and OZ-Mul, Figure 1B).

@@

This interaction model is fit for a window of time encompassing the known release event and for every other possible window in the period-of-record. The overall length of the window was treated as an adjustable hyperparameter since the length of time for an abnormal material release to reach the footprint of the flux tower may be event-dependent and/or unknown. In addition, the length of time for deposited material to induce a possible vegetation response (or at least a flux tower anomaly) is an unknown quantity. Thus, in the default base case, we set the window length (n\_days) as 7 days, but we ran hyperparameter experiments to determine the result of setting it to longer periods of 10 and 14 days. The window length affects the duration of time designated as before and after the event. The period of time designated as during the event was set as 2 days in all cases. We compared the effect size of the interaction term ($\beta$3) during the event-encompassing window against all other windows in a “rolling” analysis. Therefore, the overall analysis compares the interaction term of the window of the known event against a window centered on every other observation. Specific windows that had an effect size at least as large as that of the known event and with wind conditions defined as “towards” the flux tower were flagged as “event detections”.

@@

For the IRE case study, among all possible pairwise combinations of dependent and independent variables (n=51, Table S1), we found that only the $CO_2$ versus air temperature comparison had any measure of predictability from the model ($R^2$ > 0.05) and a strong event effect size (Table 1). Although other variable pairs such as sensible heat flux and net radiation had a stronger effect size for the BE-Lon site, they had a weak linear relationship across the other sites. When we focus on the period of the abnormal release event in particular rather than the overall period of record, we found that $CO_2$ versus air temperature relationship at each of the three focal towers had similar $R^2$ values and effect sizes (Table 1). Figure 2 provides a visual example of the effect size framing in our analysis. Note how the apparent interaction effect between time periods around the abnormal release event in Figure 2A (before, during, after) translates to a relatively high effect size (Figure 2C) whereas in Figure 2B at an arbitrary time point, there is no apparent interaction effect. This corresponds to a low effect size in Figure 2D.

@@

Another contributor to the success of our approach was the fact that Belgium has a relatively dense network of flux towers with long data records. As a result, we were able to compare event detections at the closest tower against more distant towers (refer to Figure 1), increasing our confidence that we were seeing a true signal and not spurious noise. This level of flux tower density may be found in the United States, Western/Central Europe, Japan, and Australia, which have high densities of flux towers, but not throughout the rest of the world \citep{baldocchi2001fluxnet,pastorello2020fluxnet2015}. As a result, this may impose geographical limits on our approach. Another attribute that likely aided our investigations was the fact that all three of the flux towers we examined were located in similar agricultural ecosystem types. Although investigating whether or not flux tower combinations with differing ecosystems would yield similar results is beyond the scope of the present study, we suspect that the data records among towers in disparate ecosystem types could differ too drastically to be of comparative use. For example, \citet{pastorello2020fluxnet2015} show that towers in the crop ecosystem type, of which all three sites used in this study belong, have a relatively narrow distribution of fluxes (i.e., gross primary production, GPP) compared to towers in the grassland or evergreen broadleaf forest types. One reason for this narrow spread in crop ecosystems may be the relative homogeneity of flux tower footprints in crop ecosystems, which typically extend to within 1000m of the tower depending on atmospheric conditions \citep[e.g., wind speed and direction,][]{chuRepresentativenessEddyCovarianceFlux2021}. Given that towers in other ecosystem types with a more heterogeneous flux tower footprint have a wider distribution of fluxes \citep[e.g.][]{pastorello2020fluxnet2015}, we suspect that having all the towers in our study located in a homogenous ecosystem type was helpful in reducing the spikiness and spread of the data.
